---
title: "Fitting Two-Way Factorial Designs"
author: Prof Randi Garcia, SDS 290
date: April 4, 2024
format: 
  revealjs:
    scrollable: true
    theme: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE)
library(tidyverse)
library(Stat2Data)
```

## Announcements

-   Exam 1 grades posted
-   HW 6 due tomorrow at 11:59p
-   Office hours
    -   Today: Cancelled
    -   Tomorrow: 10:50a - 12:05p
-   Where to get HW help
    -   [Spinelli center](https://www.smith.edu/qlc/tutoring.html) tutoring Sun-Thurs 7-9p, Sabin-Reed 301.
    -   Post questions to #hw6-questions channel on Slack!
    -   Direct message me on Slack.

## Agenda

1.  MSA Feedback
2.  MP2 Assignment
3.  Two-Way Factorial Design
    -   Interactions
    -   Fitting the ANOVA model

## SDS 290: How's things?

-   ""

## SDS 290 things going so well...

-   ""

## SDS 290 things not going so well

-   ""

## SDS 290 what I can do

-   

## SDS 290 what you can do

-   ""

## MP2 Assignment

-   MP2 instructions
-   MP2 [group selection survey](https://docs.google.com/forms/d/e/1FAIpQLSdAnm-YO99CPLIkkpaciS9_e0V04bCRFCVsVg0mdHAIOvSJtQ/viewform?usp=sf_link)

## Interaction Graphs

For each of the following interaction graphs, answer the following questions with **YES** or **NO**.

1.  Is there a main effect of B12?
2.  Is there a main effect of antibiotics?
3.  Is there an interaction between B12 and antibiotics?

## Practice 1

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
piglets <- data.frame(gain = c(7, 7, 7, 
                               4, 4, 4,
                               4, 4, 4,
                               7, 7, 7), 
                      group = as.factor(c(rep(1, 3), rep(2, 3), rep(3, 3), rep(4, 3))), 
                      antibiotic = c(rep("0mg", 3), rep("40mg", 3), rep("0mg", 3), rep("40mg", 3)),
                      B12 = c(rep("0mg", 6), rep("5mg", 6)))

ggplot(piglets, aes(x = B12, y = gain, 
                    group = antibiotic, 
                    linetype = antibiotic, 
                    shape = antibiotic,
                    color = antibiotic)) +
  geom_point() +
  geom_smooth(method = "lm", se = 0) +
  ylim(2, 9)
```

## Practice 1

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
piglets <- data.frame(gain = c(7, 7, 7, 
                               4, 4, 4,
                               4, 4, 4,
                               7, 7, 7), 
                      group = as.factor(c(rep(1, 3), rep(2, 3), rep(3, 3), rep(4, 3))), 
                      antibiotic = c(rep("0mg", 3), rep("40mg", 3), rep("0mg", 3), rep("40mg", 3)),
                      B12 = c(rep("0mg", 6), rep("5mg", 6)))

ggplot(piglets, aes(x = B12, y = gain, 
                    group = antibiotic, 
                    linetype = antibiotic, 
                    shape = antibiotic,
                    color = antibiotic)) +
  geom_point() +
  geom_smooth(method = "lm", se = 0) +
  ylim(2, 9)
```

1.  NO
2.  NO
3.  YES

## Practice 2

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
piglets <- data.frame(gain = c(4, 4, 4, 
                               5, 5, 5,
                               6, 6, 6,
                               7, 7, 7), 
                      group = as.factor(c(rep(1, 3), rep(2, 3), rep(3, 3), rep(4, 3))), 
                      antibiotic = c(rep("0mg", 3), rep("40mg", 3), rep("0mg", 3), rep("40mg", 3)),
                      B12 = c(rep("0mg", 6), rep("5mg", 6)))

ggplot(piglets, aes(x = B12, y = gain, 
                    group = antibiotic, 
                    linetype = antibiotic, 
                    shape = antibiotic,
                    color = antibiotic)) +
  geom_point() +
  geom_smooth(method = "lm", se = 0) +
  ylim(2, 9)
```

## Practice 2

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
piglets <- data.frame(gain = c(4, 4, 4, 
                               5, 5, 5,
                               6, 6, 6,
                               7, 7, 7), 
                      group = as.factor(c(rep(1, 3), rep(2, 3), rep(3, 3), rep(4, 3))), 
                      antibiotic = c(rep("0mg", 3), rep("40mg", 3), rep("0mg", 3), rep("40mg", 3)),
                      B12 = c(rep("0mg", 6), rep("5mg", 6)))

ggplot(piglets, aes(x = B12, y = gain, 
                    group = antibiotic, 
                    linetype = antibiotic, 
                    shape = antibiotic,
                    color = antibiotic)) +
  geom_point() +
  geom_smooth(method = "lm", se = 0) +
  ylim(2, 9)
```

1.  YES
2.  YES
3.  NO

## Practice 3

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
piglets <- data.frame(gain = c(4, 4, 4, 
                               7, 7, 7,
                               4, 4, 4,
                               7, 7, 7), 
                      group = as.factor(c(rep(1, 3), rep(2, 3), rep(3, 3), rep(4, 3))), 
                      antibiotic = c(rep("0mg", 3), rep("40mg", 3), rep("0mg", 3), rep("40mg", 3)),
                      B12 = c(rep("0mg", 6), rep("5mg", 6)))

ggplot(piglets, aes(x = B12, y = gain, 
                    group = antibiotic, 
                    linetype = antibiotic, 
                    shape = antibiotic,
                    color = antibiotic)) +
  geom_point() +
  geom_smooth(method = "lm", se = 0) +
  ylim(2, 9)
```

## Practice 3

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
piglets <- data.frame(gain = c(4, 4, 4, 
                               7, 7, 7,
                               4, 4, 4,
                               7, 7, 7), 
                      group = as.factor(c(rep(1, 3), rep(2, 3), rep(3, 3), rep(4, 3))), 
                      antibiotic = c(rep("0mg", 3), rep("40mg", 3), rep("0mg", 3), rep("40mg", 3)),
                      B12 = c(rep("0mg", 6), rep("5mg", 6)))

ggplot(piglets, aes(x = B12, y = gain, 
                    group = antibiotic, 
                    linetype = antibiotic, 
                    shape = antibiotic,
                    color = antibiotic)) +
  geom_point() +
  geom_smooth(method = "lm", se = 0) +
  ylim(2, 9)
```

1.  NO
2.  YES
3.  NO

## Two-Way Factorial Design Formulas

$${y}_{ijk}={\mu}+{\alpha}_{i}+{\beta}_{j}+{\alpha\beta}_{ij}+{e}_{ijk}$$

-   Where $i$, from 1 to $a$, is the level of the first factor,
-   $j$, from 1 to $b$, is the level of the second factor,
-   and $k$, from 1 to $n$, is the observation in each cell.

## Sum of Squares (SS)

$${SS}_{A} = \sum_{i=1}^{a}bn(\bar{y}_{i..}-\bar{y}_{…})^{2}$$

$${SS}_{B} = \sum_{j=1}^{b}an(\bar{y}_{.j.}-\bar{y}_{…})^{2}$$

$${SS}_{AB} = n\sum_{i=1}^{a}\sum_{j=1}^{b}(\bar{y}_{ij.}-\bar{y}_{i..}-\bar{y}_{.j.}+\bar{y}_{…})^{2}$$

$${SS}_{E} = \sum_{i=1}^{a}\sum_{j=1}^{b}\sum_{k=1}^{n}({y}_{ijk}-\bar{y}_{ij.})^{2}$$

## Degrees of Freedom (df)

$${df}_{A}=a-1$$

$${df}_{B}=b-1$$

$${df}_{AB}=(a-1)(b-1)$$

$${df}_{E}=ab(n-1)$$

## Mean Squares (MS)

$${MS}_{A}=\frac{{SS}_{A}}{{df}_{A}}$$

$${MS}_{B}=\frac{{SS}_{B}}{{df}_{B}}$$

$${MS}_{AB}=\frac{{SS}_{AB}}{{df}_{AB}}$$

$${MS}_{E}=\frac{{SS}_{E}}{{df}_{E}}$$

## Three Research Questions, Three F-ratios

1.  Is there a significant main effect of factor A?
2.  Is there a significant main effect of factor B?
3.  Is there an significant interaction between factor A and factor B?

## F-ratios and the F-distribution

The ultimate statistics we want to calculate is Variability in treatment effects/Variability in residuals. The F-ratio.

$$F = \frac{{MS}_{A}}{{MS}_{E}}$$

$$F = \frac{{MS}_{B}}{{MS}_{E}}$$

$$F = \frac{{MS}_{AB}}{{MS}_{E}}$$

## ANOVA Source Table for Two-Way Factorial

$${y}_{ijk}={\mu}+{\alpha}_{i}+{\beta}_{j}+{\alpha\beta}_{ij}+{e}_{ijk}$$

|         Source |                                             SS                                             |      df      |              MS               |              F               |
|--------------:|:-------------:|:-------------:|:-------------:|:-------------:|
|    Treatment A |                     $\sum_{i=1}^{a}bn(\bar{y}_{i..}-\bar{y}_{…})^{2}$                      |    $a-1$     |  $\frac{{SS}_{A}}{{df}_{A}}$  | $\frac{{MS}_{A}}{{MS}_{E}}$  |
|    Treatment B |                     $\sum_{j=1}^{b}an(\bar{y}_{.j.}-\bar{y}_{…})^{2}$                      |    $b-1$     |  $\frac{{SS}_{B}}{{df}_{B}}$  | $\frac{{MS}_{B}}{{MS}_{E}}$  |
| Interaction AB | $n\sum_{i=1}^{a}\sum_{j=1}^{b}(\bar{y}_{ij.}-\bar{y}_{i..}-\bar{y}_{.j.}+\bar{y}_{…})^{2}$ | $(a-1)(b-1)$ | $\frac{{SS}_{AB}}{{df}_{AB}}$ | $\frac{{MS}_{AB}}{{MS}_{E}}$ |
|          Error |         $\sum_{i=1}^{a}\sum_{j=1}^{b}\sum_{k=1}^{n}({y}_{ijk}-\bar{y}_{ij.})^{2}$          |  $ab(n-1)$   |  $\frac{{SS}_{E}}{{df}_{E}}$  |                              |

## Two-Way Factorial in R

[See Two-Way Factorial code](https://randilgarcia.github.io/sds290fall21/lectures/09_two-way_fac_inR.Rmd)

<!-- ## Teaching Methods -->

<!-- A professor wanted to compare three different teaching methods to determine how students  -->

<!-- would perceive the course: 1) instructionist, 2) inquiry-based, and 3) team-based. She randomly assigned the same class (same topic different students) from 6 different semesters to treatments. At the end of the semester students were asked to rate the course on a 5-point scale, and the average class rating was calculated.  -->

<!-- ## Anxiety and Memory -->

<!-- A psychologist wants to study the effect of anxiety on 4 different types of memory. Twelve participants are assigned to one of two anxiety conditions: 1) low anxiety group is told that they will be awarded \$5 for participation and \$10 if they remember sufficiently accurately, and 2) high anxiety group is told they will be awarded \$5 for participation and \$100 if they remember sufficiently accurately. All subjects perform four memory trials in random order, testing 4 different types of memory. The number of errors on each trial is recorded.      -->
